---
title: "FastDE Large Sparse Matrix Support"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FastDE Large Sparse Matrix Support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---

***

```{r, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE, echo = FALSE, results = 'hide'}
library(fastde)
library(Matrix)
```

# Handling Large Datasets
R sparse matrix, `dgCMatrix`, has a limit of 2^31 or 2 billion non-zero elements.  A data set, such as 10X Genomics' mouse brain data set with 1.3 million cells, can exceed this limit.  FastDE defines a `dgCMatrix64` class for handling large sparse matrices.

## Difference between dgCMatrix and dgCMatrix64
Both dgCMatrix and dgCMatrix64 use compressed sparse column format, represented by 3 arrays: `x` for non-zero element values, `i` for row index of each non-zero element, and `p` for offsets of each column in the `x` an `i` arrays.  `x` is a doubles array, and `i` is an integers array, as is the `Dim` array containing the extents of the 2D sparse matrix.    The `p` array in dgCMatrix64 is of type doubles, in contrast to the integer array in `dgCMatrix`.   FastDE leverages IEEE double precision floating point definition to allow integers up to 2^53 in size to be represented exactly.  This is the same approach that R uses for representing large arrays and dense matrices.  However, this change prevents dgCMatrix64 from being used as a drop-in replacement of dgCMatrix.  dgCMatrix64 currently only matrices of up to 2^31 rows or columns.

## Create a dgCMatrix64 object
A dgCMatrix64 object can be created directly by providing the `x`, `i`, `p`, and `Dim` arrays during construction.  Alternatively, an R dgCMatrix can be converted to a dgCMatrix64 object using `as.dgCMatrix64` function.

```{r create, fig.height=8, fig.width=15, eval = FALSE, echo = FALSE, results = 'hide'}
# create dgCMatrix64
spmat <- rsparsematrix(1000, 2000, 0.05)

# direct construction
spmat64 = new('dgCMatrix64', x = spmat@x, i = spmat@i, p = spmat@p, Dim = spmat@Dim, Dimnames = spmat@Dimnames)

# via as function
spmat64 <- as.dgCMatrix64(spmat)
```

## Read and write dgCMatrix64 object to files
dgCMatrix64 can be serialized to and deserialized from HDF5 files with the provided functions `Write10X_h5` and `Read10X_h5_big` functions.  Both will adapt to data size automatically to accept or product `dgCMatrix` or `dgCMatrix64` objects. 

```{r io, fig.height=8, fig.width=15, eval = FALSE, echo = FALSE, results = 'hide'}
# write 
fastde::Write10X_h5(spmat, "test_spmat.h5")

# read
spmat_2 = fastde::Read10X_h5_big("test_spmat.h5")

# write 
fastde::Write10X_h5(spmat64, "test_spmat64.h5")

# read
spmat64_2 = fastde::Read10X_h5_big("test_spmat64.h5")
```

## Supported Matrix operations
FastDE provides a few basic matrix operations, specifically, conversion to dense matrix, transpose, compute row and column sums, and concatenate matrices.  All functions support a `threads` parameter for multithread support.  Again, all functions support both dgCMatrix and dgCMatrix64.  


```{r ops, fig.height=8, fig.width=15, eval = FALSE, echo = FALSE, results = 'hide'}
# convert to dense 
mat64 = fastde::sp_to_dense(spmat64)

# transpose
spmat64_t = fastde::sp_transpose(spmat64)

# create transposed dense matrix
mat64_t = fastde::sp_to_dense_transposed(spmat64)

# row and column sums
rsums = fastde::sp_rowSums(spmat64, threads=4)
csums = fastde::sp_colSums(spmat64, threads=4)

# concatenate
spmat64_rbind = fastde::sp_rbind(list(spmat, spmat, spmat))
spmat64_cbind = fastde::sp_cbind(list(spmat, spmat, spmat))
```


## Using dgCMatrix64 for FastDE functionc
Since Seurat object by default has slot for dgCMatrix but not dgCMatrix64, FastDE provides `FastFindAllMarkers64` for `dgCMatrix64` directly.  Compared to `FastFindAllMarker`, this function requires the cell labels to be supplied as an integer array.  `FastFindMarkers` function in FastDE natively supports `dgCMatrix64` and `dgCMatrix`.  Multithrading depends on the `future::plan` configuration.

```{r findmarkers, include = FALSE, eval = FALSE, echo = FALSE, results = 'hide'}
# 
labels <- as.integer(sample(1:5, 2000, replace = TRUE))
pbmc.markers <- FastFindAllMarkers64(spmat, idents.clusters = labels, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, test.use = 'fastwmw', return.dataframe = TRUE)
```

The wilcoxon rank sum test and students t-test can also be invoked directly with either dense or sparse matrix as input.  Multithreading can be specified directly.

```{r tests, include = FALSE, eval = FALSE, echo = FALSE, results = 'hide'}
# student's t-test
res <- fastde::sparse_ttest_fast(spmat, labels, as_dataframe = FALSE, threads = 4L, alternative = 2L, var_equal = FALSE)

# wilcoxon rank sum test
fastdewilcox <- fastde::sparse_wmw_fast(spmat, labels, rtype=2L, continuity_correction=TRUE, as_dataframe = FALSE, threads = 4L)

# wilcoxon rank sum test on dense matrix
fastdewilcox <- fastde::wmw_fast(mat, labels, rtype=2L, continuity_correction=TRUE, as_dataframe = FALSE, threads = 4L)
```


```{r save.times, include = FALSE, eval = FALSE, echo = FALSE, results = 'hide'}
write.csv(x = t(as.data.frame(all_times)), file = "../output/timings/pbmc3k_tutorial_times.csv")
```
<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>